---
title: "Comparing NCBI, Silva, RAD using MetaScope"
output: html_document
author: "Remington Manning-Neal"
---

## Required packages


## Install MetaScope

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("MetaScope")
# BiocManager::install("ShortRead")
```

```{r}
suppressPackageStartupMessages({
  library(MetaScope)
  library(magrittr)
  # library(ShortRead)
})
```

## Load data

```{r}
source("label_to_organism.R")
source("load_data.R")
ncbi_db_path <- bt_indices_metascope_ncbi() # bowtie indices
silva_db_path <- bt_indices_metascope_silva() # bowtie indices
radlib16s_db_path <- bt_indices_radlib_16S() # bowtie indices
# radlibvr_db_path <- bt_indices_radlib_vr() # bowtie indices (many sequences are empty so warning suppression is needed)

dada2_testData <- load_dada2_test_data() 
v3v4_testData <- load_real_world_test_data() 

data_dir <- file.path(getwd(), "data")
ncbi_data_dir <- file.path(data_dir, "ncbi")
silva_data_dir <- file.path(data_dir, "silva")
radlib_data_dir <- file.path(data_dir, "rad")
```

## Create results directories

```{r}
analysis_dir <- file.path(getwd(), "analysis")
dir.create(analysis_dir)
results_dir <- file.path(getwd(), "results")
dir.create(results_dir)
```

# DADA2 test files

```{r}
dada2_1st_fastq_file <- list.files(path = dada2_testData, pattern = ".fastq")[1]
```

## MetaScope NCBI

### Create BAM files

```{r}
ncbi_output_temp <- tempfile(pattern = "ncbi", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(ncbi_output_temp))

ncbi_dada2_map <-
  align_target_bowtie(
    read1 = file.path(dada2_testData, dada2_1st_fastq_file),
    lib_dir = ncbi_db_path,
    libs = "ncbi",
    align_dir = ncbi_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

### Alignment Evaluation

```{r}
ncbi_dada2_bamFile <- Rsamtools::BamFile(ncbi_dada2_map)

param <-
  Rsamtools::ScanBamParam(
    flag = Rsamtools::scanBamFlag(isSecondaryAlignment = FALSE),
    what = c("flag", "rname")
  ) #Gets info about primary alignments

aln <- Rsamtools::scanBam(ncbi_dada2_bamFile, param = param)
accession_all <- aln[[1]]$rname
# genome_name_all <- accession_all |> 
#   taxonomizr::accessionToTaxa(tmp_accession) |>
#   taxonomizr::getTaxonomy(sqlFile = tmp_accession, desiredTaxa = "strain")
# read_count_table <- sort(table(genome_name_all), decreasing = TRUE)
read_count_table <- sort(table(accession_all), decreasing = TRUE)
knitr::kable(
  read_count_table[1:10],
  col.names = c("Genome Assigned", "Read Count"))
```


### MetaID (Unable to perform due to issues generating the proper taxonomizr database)

```{r}
#TODO: tmp_accessions may need to be a sqlite db. May not be needed though???
# output <- metascope_id(
#   map,
#   input_type = "bam",
#   # change input_type to "csv.gz" when not creating a BAM
#   aligner = "bowtie2",
#   num_species_plot = 0,
#   accession_path = tmp_accession
# )
# 
# knitr::kable(output,
#              format = "html",
#              digits = 2,
#              caption = "Table of MetaScope ID results")
```


## MetaScope Silva

>SILVA bowtie indices were made previously by the MetaScope team

```{r}
silva_output_temp <- tempfile(pattern = "silva", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(silva_output_temp))

silva_dada2_map <-
  align_target_bowtie(
    read1 = file.path(dada2_testData, dada2_1st_fastq_file),
    lib_dir = silva_db_path,
    libs = "silva138",
    align_dir = silva_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

### Alignment Evaluation

```{r}
silva_dada2_bamFile <- Rsamtools::BamFile(silva_dada2_map)

param <-
  Rsamtools::ScanBamParam(
    flag = Rsamtools::scanBamFlag(isSecondaryAlignment = FALSE),
    what = c("flag", "rname")
  ) #Gets info about primary alignments

aln <- Rsamtools::scanBam(silva_dada2_bamFile, param = param)
silva_dada2_accession_all <- aln[[1]]$rname
silva_dada2_read_count_table <- silva_accessions_to_organism(silva_dada2_accession_all)
# genome_name_all <- accession_all |> 
#   taxonomizr::accessionToTaxa(tmp_accession) |>
#   taxonomizr::getTaxonomy(sqlFile = tmp_accession, desiredTaxa = "strain")
# read_count_table <- sort(table(genome_name_all), decreasing = TRUE)
silva_dada2_read_count_table <- sort(table(silva_dada2_read_count_table), decreasing = TRUE)
knitr::kable(
  silva_dada2_read_count_table[1:10],
  col.names = c("Genome Assigned", "Read Count"))
```

## MetaScope RAD

```{r}
rad16_output_temp <- tempfile(pattern = "rad16", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(rad16_output_temp))

rad16_dada2_map <-
  align_target_bowtie(
    read1 = file.path(dada2_testData, dada2_1st_fastq_file),
    lib_dir = radlib16s_db_path,
    libs = "indices",
    align_dir = rad16_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

```{r}
rad16_dada2_bamFile <- Rsamtools::BamFile(rad16_dada2_map)

param <-
  Rsamtools::ScanBamParam(
    flag = Rsamtools::scanBamFlag(isSecondaryAlignment = FALSE),
    what = c("flag", "rname")
  ) #Gets info about primary alignments

aln <- Rsamtools::scanBam(rad16_dada2_bamFile, param = param)
rad_dada2_accession_all <- aln[[1]]$rname
# genome_name_all <- accession_all |> 
#   taxonomizr::accessionToTaxa(tmp_accession) |>
#   taxonomizr::getTaxonomy(sqlFile = tmp_accession, desiredTaxa = "strain")
# read_count_table <- sort(table(genome_name_all), decreasing = TRUE)
rad_dada2_read_count_table <- sort(table(rad_dada2_accession_all), decreasing = TRUE)
knitr::kable(
  rad_dada2_read_count_table[1:10],
  col.names = c("Genome Assigned", "Read Count"))
```

## BLAST baseline

```{r}
# dada2_fasta_version <- file.path(getwd(), "data", "test-data", "dada2_test_data.fasta")
# writeFasta(dada2_1st_fastq_file, dada2_fasta_version)
# 
# args = paste0("-db nt -query ", dada2_fasta_version, " -remote")
# blast_results = system2("blastn", args=args, stdout = T)

```

# V3V4 test files

```{r}
v3v4_1st_fastq_file <- list.files(path = v3v4_testData, pattern = ".fastq")[2]
```

## MetaScope NCBI

```{r}
ncbi_output_temp <- tempfile(pattern = "ncbi", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(ncbi_output_temp))

ncbi_v3v4_map <-
  align_target_bowtie(
    read1 = file.path(v3v4_testData, v3v4_1st_fastq_file),
    lib_dir = ncbi_db_path,
    libs = "ncbi",
    align_dir = ncbi_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

### MetaID

```{r}
#TODO: tmp_accessions may need to be a sqlite db. May not be needed though???
# output <- metascope_id(
#   map,
#   input_type = "bam",
#   # change input_type to "csv.gz" when not creating a BAM
#   aligner = "bowtie2",
#   num_species_plot = 0,
#   accession_path = tmp_accession
# )
# 
# knitr::kable(output,
#              format = "html",
#              digits = 2,
#              caption = "Table of MetaScope ID results")
```


## MetaScope Silva

>SILVA bowtie indices were made previously by the MetaScope team

```{r}
silva_output_temp <- tempfile(pattern = "silva", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(silva_output_temp))

silva_v3v4_map <-
  align_target_bowtie(
    read1 = file.path(v3v4_testData, v3v4_1st_fastq_file),
    lib_dir = silva_db_path,
    libs = "silva138",
    align_dir = silva_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

## MetaScope RAD

```{r}
rad16s_output_temp <- tempfile(pattern = "rad16s", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(rad16s_output_temp))

rad16_v3v4_map <-
  align_target_bowtie(
    read1 = file.path(v3v4_testData, v3v4_1st_fastq_file),
    lib_dir = radlib16s_db_path,
    libs = "indices",
    align_dir = rad16s_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

## BLAST baseline

```{r}

```

## Analysis

```{r}

```

## Figure

```{r}

```

