---
title: "Comparing NCBI, Silva, RAD using MetaScope"
output: html_document
author: "Remington Manning-Neal"
---

## Required packages


## Install MetaScope

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("MetaScope")
# BiocManager::install("ShortRead")
```

```{r}
suppressPackageStartupMessages({
  library(MetaScope)
  library(magrittr)
  # library(ShortRead)
})
library(data.table)
library(tidyverse)
library(ggtext)
```

## Load data

```{r}
source("label_to_organism.R")
source("load_data.R")
ncbi_db_path <- bt_indices_metascope_ncbi() # bowtie indices
silva_db_path <- bt_indices_metascope_silva() # bowtie indices
radlib16s_db_path <- bt_indices_radlib_16S() # bowtie indices
# radlibvr_db_path <- bt_indices_radlib_vr() # bowtie indices (many sequences are empty so warning suppression is needed)

dada2_testData <- load_dada2_test_data() 
v3v4_testData <- load_real_world_test_data() 

data_dir <- file.path(getwd(), "data")
ncbi_data_dir <- file.path(data_dir, "ncbi")
silva_data_dir <- file.path(data_dir, "silva")
radlib_data_dir <- file.path(data_dir, "rad")
```

## Create results directories

```{r}
analysis_dir <- file.path(getwd(), "analysis")
dir.create(analysis_dir)
results_dir <- file.path(getwd(), "results")
dir.create(results_dir)
```

# DADA2 test files

```{r}
dada2_1st_fastq_file <- list.files(path = dada2_testData, pattern = ".fastq")[1]
```

## MetaScope NCBI

### Create BAM files

```{r}
ncbi_output_temp <- tempfile(pattern = "ncbi", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(ncbi_output_temp))

ncbi_dada2_map <-
  align_target_bowtie(
    read1 = file.path(dada2_testData, dada2_1st_fastq_file),
    lib_dir = ncbi_db_path,
    libs = "ncbi",
    align_dir = ncbi_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

### Alignment Evaluation

```{r}
ncbi_dada2_bamFile <- Rsamtools::BamFile(ncbi_dada2_map)

param <-
  Rsamtools::ScanBamParam(
    flag = Rsamtools::scanBamFlag(isSecondaryAlignment = FALSE),
    what = c("flag", "rname")
  ) #Gets info about primary alignments

aln <- Rsamtools::scanBam(ncbi_dada2_bamFile, param = param)
accession_all <- aln[[1]]$rname
# genome_name_all <- accession_all |> 
#   taxonomizr::accessionToTaxa(tmp_accession) |>
#   taxonomizr::getTaxonomy(sqlFile = tmp_accession, desiredTaxa = "strain")
# read_count_table <- sort(table(genome_name_all), decreasing = TRUE)
ncbi_dada2_read_count_table <- sort(table(accession_all), decreasing = TRUE)
knitr::kable(
  ncbi_dada2_read_count_table[1:10],
  col.names = c("Genome Assigned", "Read Count"))
```

## MetaScope Silva

>SILVA bowtie indices were made previously by the MetaScope team

```{r}
silva_output_temp <- tempfile(pattern = "silva", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(silva_output_temp))

silva_dada2_map <-
  align_target_bowtie(
    read1 = file.path(dada2_testData, dada2_1st_fastq_file),
    lib_dir = silva_db_path,
    libs = "silva138",
    align_dir = silva_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

### Alignment Evaluation

```{r}
silva_dada2_bamFile <- Rsamtools::BamFile(silva_dada2_map)

param <-
  Rsamtools::ScanBamParam(
    flag = Rsamtools::scanBamFlag(isSecondaryAlignment = FALSE),
    what = c("flag", "rname")
  ) #Gets info about primary alignments

aln <- Rsamtools::scanBam(silva_dada2_bamFile, param = param)
silva_dada2_accession_all <- aln[[1]]$rname
silva_dada2_read_count_table <- silva_accessions_to_organism(silva_dada2_accession_all)
# genome_name_all <- accession_all |> 
#   taxonomizr::accessionToTaxa(tmp_accession) |>
#   taxonomizr::getTaxonomy(sqlFile = tmp_accession, desiredTaxa = "strain")
# read_count_table <- sort(table(genome_name_all), decreasing = TRUE)
silva_dada2_read_count_table <- sort(table(silva_dada2_read_count_table), decreasing = TRUE)
knitr::kable(
  silva_dada2_read_count_table[1:10],
  col.names = c("Genome Assigned", "Read Count"))
```

## MetaScope RAD

```{r}
rad16_output_temp <- tempfile(pattern = "rad16", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(rad16_output_temp))

rad16_dada2_map <-
  align_target_bowtie(
    read1 = file.path(dada2_testData, dada2_1st_fastq_file),
    lib_dir = radlib16s_db_path,
    libs = "indices",
    align_dir = rad16_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

```{r}
rad16_dada2_bamFile <- Rsamtools::BamFile(rad16_dada2_map)

param <-
  Rsamtools::ScanBamParam(
    flag = Rsamtools::scanBamFlag(isSecondaryAlignment = FALSE),
    what = c("flag", "rname")
  ) #Gets info about primary alignments

aln <- Rsamtools::scanBam(rad16_dada2_bamFile, param = param)
rad_dada2_accession_all <- aln[[1]]$rname
# genome_name_all <- accession_all |> 
#   taxonomizr::accessionToTaxa(tmp_accession) |>
#   taxonomizr::getTaxonomy(sqlFile = tmp_accession, desiredTaxa = "strain")
# read_count_table <- sort(table(genome_name_all), decreasing = TRUE)
rad_dada2_read_count_table <- sort(table(rad_dada2_accession_all), decreasing = TRUE)
knitr::kable(
  rad_dada2_read_count_table[1:10],
  col.names = c("Genome Assigned", "Read Count"))
```

## BLAST baseline

```{r}
# dada2_fasta_version <- file.path(getwd(), "data", "test-data", "dada2_test_data.fasta")
# writeFasta(dada2_1st_fastq_file, dada2_fasta_version)
# 
# args = paste0("-db nt -query ", dada2_fasta_version, " -remote")
# blast_results = system2("blastn", args=args, stdout = T)

```

# V3V4 test files

```{r}
v3v4_1st_fastq_file <- list.files(path = v3v4_testData, pattern = ".fastq")[2]
```

## MetaScope NCBI

```{r}
ncbi_output_temp <- tempfile(pattern = "ncbi", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(ncbi_output_temp))

ncbi_v3v4_map <-
  align_target_bowtie(
    read1 = file.path(v3v4_testData, v3v4_1st_fastq_file),
    lib_dir = ncbi_db_path,
    libs = "ncbi",
    align_dir = ncbi_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

## MetaScope Silva

>SILVA bowtie indices were made previously by the MetaScope team

```{r}
silva_output_temp <- tempfile(pattern = "silva", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(silva_output_temp))

silva_v3v4_map <-
  align_target_bowtie(
    read1 = file.path(v3v4_testData, v3v4_1st_fastq_file),
    lib_dir = silva_db_path,
    libs = "silva138",
    align_dir = silva_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

## MetaScope RAD

```{r}
rad16s_output_temp <- tempfile(pattern = "rad16s", tmpdir = file.path(getwd(), "analysis"))
dir.create(file.path(rad16s_output_temp))

rad16_v3v4_map <-
  align_target_bowtie(
    read1 = file.path(v3v4_testData, v3v4_1st_fastq_file),
    lib_dir = radlib16s_db_path,
    libs = "indices",
    align_dir = rad16s_output_temp,
    align_file = "bowtie_target",
    overwrite = TRUE
  )
```

## BLAST baseline

```{r}

```

## Analysis

```{r}

```

## Figure

```{r}
get_fig_counts <- function() {
  # ncbi_counts  <- read_csv("figures/figure-3/data/ncbi_counts_with_names.csv")
  # rad_counts   <- read_csv("figures/figure-3/data/rad_counts_with_names.csv")
  # silva_counts <- read_csv("figures/figure-3/data/silva_counts_with_names.csv")
  ncbi_counts <- ncbi_accessions_to_organism(ncbi_dada2_read_count_table)
  rad_counts <- rad_accessions_to_organism(rad_dada2_read_count_table)
  silva_counts <- silva_accessions_to_organism(silva_dada2_read_count_table)

  ncbi_species  <- ncbi_counts  %>% filter(Freq > 0) %>% pull(organism_short) %>% unique()
  rad_species   <- rad_counts   %>% filter(Freq > 0) %>% pull(organism_short) %>% unique()
  silva_species <- silva_counts %>% filter(Freq > 0) %>% pull(organism_short) %>% unique()

  common_species <- Reduce(intersect, list(ncbi_species, rad_species, silva_species))
  common_species <- common_species[common_species != "Bacteroides caecimuris"]

  combined <- bind_rows(
    ncbi_counts  %>% mutate(database = "NCBI - MetaScope"),
    rad_counts   %>% mutate(database = "RADlib - MetaScope"),
    silva_counts %>% mutate(database = "SILVA - MetaScope")
  ) %>%
    filter(organism_short %in% common_species, Freq > 0) %>%
    group_by(organism_short, database) %>%
    summarise(Freq = sum(Freq), .groups = "drop")

  return(combined)
}

# temporary fake blast data
add_blast_fake_data <- function(combined) {
  blast_fake_data <- combined %>%
    distinct(organism_short) %>%
    mutate(database = "NCBI - BLAST", Freq = 0.1)

  return(bind_rows(combined, blast_fake_data))
}

# add in italics for Genus species names
add_labels <- function(combined) {
  combined %>%
    mutate(label = paste0("<i>", word(organism_short, 1), " ", word(organism_short, 2), "</i>"))
}

# main figure generation code
generate_figure <- function() {
  combined <- get_fig_counts() %>%
    add_blast_fake_data() %>%
    add_labels()

  ggplot(combined, aes(x = Freq, y = label, fill = database)) +
    geom_col(position = position_dodge(0.9), alpha = 0.9,
             color = "black", linewidth = 0.3) +
    scale_fill_manual(
      values = c(
        "NCBI - BLAST"       = "#E8871A",
        "NCBI - MetaScope"   = "#888780",
        "SILVA - MetaScope"  = "#378ADD",
        "RADlib - MetaScope" = "#1D9E75"
      ),
      breaks = c(
        "NCBI - BLAST",
        "NCBI - MetaScope",
        "SILVA - MetaScope",
        "RADlib - MetaScope"
      )
    ) +
    labs(
      x    = "Read count",
      y    = NULL,
      fill = "Database - Algorithm"
    ) +
    theme_bw(base_size = 15) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_line(color = "grey90"),
      axis.text.y        = element_markdown(size = 15),
      legend.position    = "right",
      legend.key.size    = unit(0.8, "cm")
    )
}

generate_figure()
```

